---
title: "Contingency tables and chisq tests"
author: "Wing-Ho Ko"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
```

### Contingency tables

#### Converting data to contingency table

```{r}
gss_sim <- gss_cat |> 
  filter(year==2014 & race != "Not Applicable" & marital != "No answer") |> 
  mutate(
    marital = marital |> fct_recode("Never" = "Never married") |>
      fct_drop("No answer"),
    race = fct_drop(race, "Not applicable")
  )
```


```{r}
glimpse(gss_sim)
```

Creates cross tabs using `table()`

```{r}
table(gss_sim$race, gss_sim$marital)
```

Do the same using xtabs, which has a formula interface

```{r}
xtabs(~ race + marital, gss_sim)
```

Suppose we start with count data already:

```{r}
gss_summary <- gss_sim |> group_by(race, marital) |> count() |> ungroup()
glimpse(gss_summary)
```

In which case the `xtab` function can still be used to produce crosstab, but the formula has to include the column of count on the LHS

```{r}
xtabs(n ~ race + marital, gss_summary)
```

#### Adding marginal counts; convert to frequency table

Save the previous generated crosstab into a variable

```{r}
gss_xtab2 <- xtabs(n ~ race + marital, gss_summary)
```

Create a crosstab with marginal counts

```{r}
addmargins(gss_xtab2)
```

Convert the crosstabs of count into crosstab of frequency

```{r}
proportions(gss_xtab2)
```

### Chi-square goodness-of-fit test

Construct the count for the race of respondents:

```{r}
gss_xtab1 <- xtabs(~ race, gss_sim)
print(gss_xtab1)
```

We'll check if the proportion of respondents from this sample is representative of the race proportion found in the the 2020 census, where 71.0% of people report to be White (alone or in combination), and 14.2% of people report to be Black (alone or in combination):

```{r}
race_census <- c(0.148, 0.142, 0.710)
chisq.test(gss_xtab1, p = race_census, rescale.p = TRUE)
```

### Pearson's Chi-square test

Return to the two-dimensional crosstab previously computed

```{r}
gss_xtab2
```

We'll now check if there's any interaction between race and martial status:

```{r}
chisq.test(gss_xtab2)
```

Since the p value is small, we conclude that there are interactions between race and martial status.
