---
title: "t-test demo"
author: "Wing-Ho Ko"
output: html_notebook
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(effectsize)
```


## t-test demos

### Orca example (simulated data)

Suppose we have tagged 25 orcas in the Washington shore and we recorded their weights one year apart. The data is stored in the `orca_data.csv` file in the data folder

Perform the appropriate t-test to check if there’s discernible difference between the weights of orcas between the two measurement

**t-test to perform**: the observations are paired. Thus a paired t-test is needed

Load the data and inspect it

```{r}
orcas <- read_csv("data/orca_simulated.csv")
glimpse(orcas)
```

Subset data that can be used as input to t.test()

```{r}
orcas_1 <- orcas |> filter(year==2023) |> arrange(tag)
orcas_2 <- orcas |> filter(year==2024) |> arrange(tag)

orcas_wt_1 <- orcas_1$weight_kg
orcas_wt_2 <- orcas_2$weight_kg
```

Perform t.test. Defaulted to printing out the results

```{r}
t.test(orcas_wt_2, orcas_wt_1, alternative="two.sided", paired=TRUE)
```

Assign the `t.test()` results to variable. Extract the p-value programatically:

```{r}
orcas_t <- t.test(orcas_wt_2, orcas_wt_1, alternative="two.sided", paired=TRUE)
```

Extract the p-value programatically:

```{r}
print(orcas_t$p.value)
```

Extract the point estimate and confidence interval programatically:

```{r}
print(orcas_t$estimate)
```

```{r}
print(orcas_t$conf.int)
```

## t-test revisited

Run different t-tests against the orca data

```{r}
orcas_t1 <- t.test(orcas_wt_2, orcas_wt_1, alternative="two.sided", paired=TRUE)
orcas_t2 <- t.test(orcas_wt_2, orcas_wt_1, alternative="two.sided", var.equal=TRUE)
orcas_t3 <- t.test(orcas_wt_2, orcas_wt_1, alternative="two.sided")
```

Print the different p-values obtained. Observed that p-values become worse with paired vs (unpaired pooled) vs (unpaired unpooled)

```{r}
print(c(orcas_t1$p.value, orcas_t2$p.value, orcas_t3$p.value))
```

### Statistical power

Check the statistical power assuming that the difference is "real"

```{r}
orcas_diff <- mean(orcas_wt_1) - mean(orcas_wt_2)
orcas_diff_sd <- sd(orcas_wt_1 - orcas_wt_2)

power.t.test(n=20, delta=orcas_diff, sd=orcas_diff_sd, sig.level=0.05, type="paired")
```

If instead we use an unpaired, pooled t-test:

```{r}
orcas_diff <- mean(orcas_wt_1) - mean(orcas_wt_2)
orcas_sd <- sqrt(0.5  * (sd(orcas_wt_1)^2 + sd(orcas_wt_2)^2))

power.t.test(n=20, delta=orcas_diff, sd=orcas_sd, sig.level=0.05)
```

Asking the question differently: how big does `n` needs to be for power = 0.80?

```{r}
power.t.test(
  power=0.80, delta=orcas_diff, sd=orcas_diff_sd, sig.level=0.05, 
  type="paired"
)
```

### Shapiro-Wilk test for normality

Use the water temperature data from Port Townsend buoy, year 2015 and 2024

```{r}
PT2015 <- read_csv("data/Port_Townsend_buoy_2015.csv")
PT2024 <- read_csv("data/Port_Townsend_buoy_2024.csv")
```

```{r}
PT2024
```


Check if either temperature data is normally distributed

```{r}
shapiro.test(PT2015$WTMP)
```

```{r}
shapiro.test(PT2024$WTMP)
```

So we shouldn't use t test to compare the two.

However, if we focus only on June,

```{r}
PT2015_06 <- filter(PT2015, month(date)==6)
shapiro.test(PT2015_06$WTMP)
```

```{r}
PT2024_06 <- filter(PT2024, month(date)==6)
shapiro.test(PT2024_06$WTMP)
```

Then it may be appropriate to perform t-test as comparison.

### F test for equal variance

Now we check if the June data from the two years have roughly the same variance

```{r}
var.test(PT2024_06$WTMP, PT2015_06$WTMP)
```

So using should use Welch's t-test

### Effect size

Now compare the 2020 and 2024 June data using t-test

```{r}
t.test(PT2024_06$WTMP, PT2015_06$WTMP)
```

So the difference is statistically discernible. Is it realistically important?

```{r}
cohens_d(PT2024_06$WTMP, PT2015_06$WTMP, pooled=FALSE)
```

We have a cohen's d that is between 1 and 2, so it the difference is important in real terms

### Wilcoxon test

Now go back to the year-long data. We'll need to use the Wilcoxon test instead:

```{r}
wilcox.test(PT2015$WTMP, PT2024$WTMP)
```

So the change in mean sea temperature is statistically significant.
