---
title: "Week 4 exercises"
author: "Wing-Ho Ko"
output: html_notebook
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(readr)
library(tidyr)
library(dplyr)
library(effectsize)
library(palmerpenguins)
```

### Concept check 1

1. Suppose you reject the null hypothesis in a one-tailed test. Will you reject the corresponding null hypothesis in a two-tailed test, at the same significance level α?

2. You are trying to determine if there’s real difference in the mean between two populations A and B. Suppose the difference is real, by increasing the sample size, would you expect…

 a. Higher statistical power?
 b. Higher effect size?
 c. Lower p-value?

### Exercise 2

Data on seawater at near-surface depth (d < 5 m) were taken at 3 different locations in the Salish Sea. The data can be found in the file "salish_cruise_simulated_1.csv". Perform an appropriate t-test on the difference in mean salinity between the Admiralty Inset samples and the Main Basin samples

### Exercise 3

Going back to the Salish Sea water samples in exercise #2. Determine if t-test was appropriated, and if so, whether it is OK to pool the variance

### Exercise 4

More Salish Sea data seawater data were taken at the 3 locations as in example 1, and are presented in the "salish_cruise_simulated_2.csv" file. This time, each set of data was collected over a broader range of location. 

1. Check if t-test is still appropriate for these samples and perform an appropriate test on the difference in mean salinity, again between Tacoma Narrows samples and the Main Basin

2. In addition, compute the effect size of the observed difference

### Exercise 5

Consider the penguins dataset.

1. Test the hypothesis that the main and female penguins are of equal proportion

2. Construct a frequency cross-tab of sex by species, with marginal sums

3. Check the hypothesis that sex ratio is independent of species
